---
title: "Gulfstream AIS"
output: html_document
date: "2025-09-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(terra)
library(sf)
library(lubridate)

```


Data were pulled from the [NOAA Marine Cadastre hub](https://hub.marinecadastre.gov/pages/vesseltraffic), which provides up-to-date AIS data for the U.S. in various forms. We only needed annual vessel counts by cell in a small region so the bulk download option was sufficient. Vessel tracks data provides locations, vessel types, and dates we can use to aggregate by month.

```{r AISdat}

gsExt<-ext(-84,-78,23,27) ##(xmin, xmax, ymin, ymax)

```

### Crop large dataset to Gulf Stream area

```{r}

### VESSEL TRACKS DATA
# vtd<-st_read("data/AISVesselTracks2021.gpkg") ### change for each year
# vtd<-st_read(dsn = "data/AISVesselTracks2019.gdb")
 
 
# gsBB<-st_bbox(c(xmin=-84,xmax=-78,ymin=23,ymax=27),crs = crs(vtd)) ##(xmin, xmax, ymin, ymax)

# now()
# # # vtdFl<-st_intersection(vtd,st_as_sfc(gsBB))
# intsct<-st_intersects(vtd,st_as_sfc(gsBB))
# vtdFl <- vtd[lengths(intsct) > 0,]
# now() # 2023 & 2022 data took ~50min-1hr to complete the intersects call
# write_sf(vtdFl,"outputs/gsTracks/aisTracksGskeys19.shp") ### change for each year

```

### Pull in inputs

```{r}

vtdFl<-st_read("outputs/gsTracks/aisTracksGskeys19.shp")%>% ### change for each year
  mutate(month=month(TrckStT))%>%
  select(month,VsslGrp)

# vtdFl$month<-month(vtdFl$TrckStT)

### create empty raster with same geometry as the data

gsr<-rast(ext=gsExt,crs=crs(vtdFl),resolution=0.001)
# gsr<-rast(ext=gsExt,crs=crs(vtdFl),resolution=0.01)

### separate df into separate by vessel group

# vGrps<-unique(vtdFl$VsslGrp)
vGrps<-c("Cargo","Fishing","Other","Passenger","Pleasure Craft/Sailing","Tanker","TugTow")

```

### rasterize tracks by month (j) and type (i)

```{r}
start<-now()
for (i in 1:length(vGrps)){ #
  
  for(j in 1:12){
  
  subDf<-vtdFl%>%
    filter(VsslGrp==vGrps[i],
           month==j)
  
  tmpR<-rasterizeGeom(vect(subDf),gsr,fun="count")
  
  # assign(paste0("rast",i),tmpR)
  if(j==1){
    monStk<-tmpR
    } else {
      monStk<-c(monStk,tmpR)
  }
  
  print(j)
  
  }
  
  print(vGrps[i])
  
  if(i==1){
    typStk<-mean(monStk)
    } else {
      typStk<-c(typStk,mean(monStk))
      }
  
}

end_time <- now()
run_time <- end_time - start

###
names(typStk)<-vGrps
# writeRaster(typStk,"outputs/aisOut/aisMonAveType2019.tif",overwrite=TRUE)

```

### Take average monthly across years and type

```{r}

ais24<-rast("outputs/aisOut/aisMonAveType2024.tif")
ais23<-rast("outputs/aisOut/aisMonAveType2023.tif")
ais22<-rast("outputs/aisOut/aisMonAveType2022.tif")
ais21<-rast("outputs/aisOut/aisMonAveType2021.tif")
ais19<-rast("outputs/aisOut/aisMonAveType2019.tif")


### Pull in monthly aves to calculate monthly aves over all 5 years

for(i in 1:length(vGrps)){
  
  vtStk<-c(ais24[[i]],ais23[[i]],ais22[[i]],ais21[[i]],ais19[[i]])
  
  
  if(i==1){
    outStk<-mean(vtStk)
    } else {
      outStk<-c(outStk,mean(vtStk))
      }
  
}

names(outStk)<-names(ais24)
# writeRaster(outStk,"outputs/aisOut/aisCountsByType.tif")


```

